---
title: "Basic Programming in R: Introduction to Programming "
output:
  pdf_document:
    keep_tex: false
    latex_engine: pdflatex
    template: ../../latex_template/readable.tex
author:
  - name: John Novembre
    affiliation: University of Chicago
date: August 26, 2026
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
spacing: single
graphics: yes
endnote: no
chapter: 1
thanks: "This document is included as part of the 'Basic Programming in R' tutorial packet for the BSD qBio Bootcamp, University of Chicago, 2026. **Current version**: `r format(Sys.time(), '%B %d, %Y')`."
abstract: "**Goal:** Introduce students to key concepts of programming in the context of the programming language `R`.   **Audience:** Incoming graduate students with no or minimal exposure to programming."
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=4, fig.path='figs/',
                      echo=TRUE, eval=FALSE, warning=FALSE, message=FALSE)

library(dplyr)
library(ggplot2)
library(reshape2)
library(tibble)
```


<!--
Internal notes 

We'll begin with an introductory section where we discuss the idea of an algorithms. We'll emphasize how to do reproducible science. We need to design algorithms that will reproduce our work.  We'll talk about key elements of algorithms, such as iterative loops that are implemented with FOR or WHILE statements. We'll discuss the idea of conditioning on variables using if-then statements (maybe introduce data types at this point too). We'll discuss the idea of modular pieces of code that can be wrapped up into functions. And we'll introduce the idea of code libraries, collections of functions that we can use.  

Warm-up Need to introduce notion of variables and state of environment; introduce Rstudio; 

As one key activity, we'll deal with a canonical situation that arises for researchers in which they receive a table of data and have to iterate over the rows, performing some analysis on each row. (Note data types while introducing table).  We'll use for loops and if-then statements as a basic beginning to do this. We'll then also show off some shortcut functions/data structures in R that allow one to do things compactly (ok to use tibbles perhaps). For the data table, I'm thinking of using a data table of perhaps RNA expression levels across genes.  Compute p-val of Xuanyao's multiple testing.   

Then, as a second level, we will have the canonical situation of having multiple files where we need to iterate over files and perform an analysis. For the files, I'm thinking about using a collection of FastA or FastQ files from a sequencing experiment. 

As a final exercise, we'll talk about the idea of simulation and simulate random walks in preparation for the defensive programming. 
-->


### Preliminaries

Please use git/gitkraken to pull the latest version of the course materials from https://github.com/jnovembre/BSD-QBio 

### Motivation 

There are several key reasons that motivate us to learn programming in the context of doing research in modern biology. The first is that it allows us more precise control over what the computer is doing and allows us to build custom solutions to our data handling and analysis problems. The second is that it allows us to automate things so that we can do simple tasks over and over and over again very efficiently. The third is in terms of reproducible science. The more that we can automate and build software workflows with scripts that will execute our analysis, the easier it is for us to reproduce the analysis that we've done before and to share them with others. We also find it extremely convenient to change a setting on our analysis and re-run it from start to finish to test the robustness of our work, and that ultimately results in better science. 

Two key concepts that are important to understand in programming are algorithms and data structures. Here I'm going to focus on introducing algorithms, and we'll pick up knowledge about data structures along the way. 

### Algorithms

An algorithm is a complex set of instructions that can be followed in an automatic way by an actor or agent, or in this case, our computer. Algorithms are everywhere in some sense. Consider the need to walk to work when there is a possibility of rain. Most of us execute a basic algorithm that looks something like the following: The first step involves checking the weather. If there's a forecast with potential rain, then grab the umbrella. If there's a forecast for clear skies, then do not grab the umbrella.  Then we head to work, and we repeat this every work-day.

A key concept we'll use here and in other parts of the Boot Camp is the idea of pseudo-code, where we write out in human-readable language what our algorithm is. 

At a high level here is pseudo-code for our walk-to-work algorithm:

1) Check the weather
2) If there is a forecast for rain, grab umbrella
3) Otherwise, leave the umbrella. 
4) Head out the door
5) If it is raining, deploy umbrella
6) else, leave it closed. 
7) Repeat every weekday morning

This simple example already brings up the idea of a key concept in programming: the conditional if/then statement. The general format for these looks like "if (boolean statement) then (action)".  A "boolean" statement is one that evaluates as true/false (e.g. "there is a forecast for rain" is true or false; "it is raining" is true or false.  The last line, about repeat every weekday morning, brings up the idea of what we call a loop. A loop is where our instructions repeat themselves a certain number of times or over a certain set of values.  

Notice how, in writing out the pseudo code, I've added steps that weren't in my original description of the problem. That is an example of a key idea that I want to express about the importance of writing how programs and thinking programmatically. By taking the time to write out even the pseudocode of the algorithm, it improved our thinking and helped us catch details that we missed when we first thought about the problem.   

Let's see these ideas in R.  First, let's do the simplest program possible, in which we print a message to the screen.  We will write our code in small code blocks, and execute them using the green arrow.   

```{r}
# First let's do the simplest program possible
print("hello world")  # Print a custom message
```

Now I'll introduce the idea of abstract variables. Much like in math, we can assign values to abstract variables like x here, and then we can print a message that tells us what x equals.   
```{r}

x <- 5 # Declaring a simple variable x and assign it value 5 
print(paste("x equals ",x))  # paste joins the variable x with a message
```

And here we demonstrate an if-then statement. Let's re-run this multiple times with multiple values. 
```{r}

x <- 5 # Declaring a simple variable x and assign it value 5 

# example of an if-then statement
if (x > 5) { 
  print ("Wow  greater than 5!") 
} else { 
    print ("Number is less than or equal to 5")
}
```


Now let's consider a short program that takes in an array of forecast values and prints out whether we should bring an umbrella or not for each day. 
```{r}

# A "vector"/array of forecast values.  Each is a "character string"
forecasts <- c("rainy","clear","rainy","snow","clear")

# We can access array values by "index".  Here are some examples...
print("==Test prints==")
print(forecasts[2])
print(forecasts[1])

# Now let's put this together, to loop over the forecast days, and print a recommendation...
print("==Plan for the week==")
for (i in 1:5){
  if(forecasts[i]=="rainy"){
    print(paste("Day",i,"bring umbrella!"))
  } else {
    print(paste("Day",i,"no forecast of rain."))
  }
}
```

### Consider Scratch as a programming language

Visit: https://scratch.mit.edu/projects/editor/?tutorial=getStarted

Use the browser on the left to explore the possible parts of a Scratch program.  Exercise: Make a short program using at least one motion, one conditional control feature (if/then), one loop control feature (e.g. wait/repeat) and one variable. For example, have a character speak out aloud 1 to 10 but it will stop and make a sound if the mouse is clicked.  


### A typical challenge case in biological research

A typical challenge case in biological research is that we have a large data table that we would like to process and work with. We oftentimes want to do some kind of processing on each row and then loop over rows. 

In this case, we're going to work on data from a study of patients with inflamed versus non-inflamed biopsies. The investigators sought to understand gene regulatory differences between these two tissue states, so they conducted RNA-seq experiments to measure the gene expression levels. We start with their data table of gene expression results.

To investigate the data, we're going to put our if-then conditionals and for loops to work. 

But first, let's read in the file.  R has a convenient function `read.csv` for reading in comma-separated value text. It also has the `head` command that's useful for looking at the top of a file. We can also use the R studio interface to look at the data. 

```{r}
x<-read.csv("Table_DEGs_GSE193677.csv")
head(x)
```
To introduce the fields (i.e. columns here), we have GeneSymbol, GeneName, logFC (which means log fold change, meaning how much did the gene expression change, but on a log scale), AveExpr (which is average expression level over both conditions), t.stat (which is a t-statistic for a test of difference in expression levels), p.value (which is the p-value for a statistical test of difference), adj.p.val (which is a p-value adjusted for mulitple testing), and "Regulation" (Up/Down).

As a first pass, let's write a simple script that counts the number of up-regulated genes:

```{r}
dim(x) # gives us dimensions of the table
ngenes <- dim(x)[1]

ctr <- 0
for(i in 1:ngenes){
  if(x$Regulation[i] == "Up"){
    ctr <- ctr + 1
  }
}
print(paste("The number that appear up-regulated is ",ctr))
```

Notice how this second code block uses x, which was defined in the previous code block. That's because in our studio we have persistent memory over code blocks. That is to say that when we declare a variable in one block and assign it a value, the existence of that variable and its contents continue until we reset the memory manually.  The RStudio interface has the notion of an environment explorer on the upper right, where we can directly inspect variables that we've declared.   

### Data types

One thing we've glossed over so far, every value in R has a *type* that determines what operations make sense on it. The four you'll encounter most:

- **numeric**: real numbers, like `3.14` or `-2.7`  
- **character**: text in quotes, like `"TP53"` or `"rainy"`  
- **logical**: exactly `TRUE` or `FALSE`  
- **integer**: whole numbers stored precisely, written `42L`

You can inspect the type of any variable with `class()`:

```{r}
gene_count    <- 487         # numeric
gene_name     <- "TP53"      # character
is_expressed  <- TRUE        # logical

class(gene_count)
class(gene_name)
class(is_expressed)
```
Types matter because R enforces them. Try this:

```{r}
gene_count + 10       # fine: arithmetic on a numeric
#gene_count + gene_name  # uncomment to see the error
```

Notice that x > 5 from our earlier example produces a logical value — that's what the if statement actually tests. Types are everywhere, even when invisible.

```{r}
x <- 7
result <- x > 5      # result is TRUE or FALSE
class(result)
print(result)
```

You'll see types come back in the Defensive Programming tutorial, where you'll write code that explicitly checks its inputs are the right type before proceeding.


### Activity 

Now I want to think of another query that you can make on each row of the RNAseq data table and then implement it in your own code block mimick'ing the code block above. Be prepared to post your code to slack when you finish. 


### Example of simplifications available in R 

Well, almost every programming language has the notion of loops and conditional statements. There are a number of built-in simplifications available in R for working with data tables because data tables are so common in statistical analysis. This is an example of how a specialized language like R can be very useful. Uh. 

Note also, rather than running this as a single program, let's run each of these commands in the console. We can highlight a line and then hit Command Enter to run just that line and inspect the output in the console. This is a more interactive way of doing data analysis than writing small programs. But ultimately, it's less reproducible, and thus less ideal, and it should be reserved for debugging and exploration. 

```{r}

x<-read.csv("Table_DEGs_GSE193677.csv")

## Count all the cases where the gene is up-regulated
sum(x$Regulation == "Up")

# Examine max values and find the max logFC gene
max(x$logFC)
which.max(x$logFC)
x$GeneSymbol[which.max(x$logFC)]

# Compute a table of counts for all unique values of a "factor"/categorical variable 
table(x$Regulation)

# Plot a histogram of the p-values
hist(x$p.value)

# Make a simple plot
plot(x$logFC,-log10(x$adj.p.val),col=(x$adj.p.val < 0.05)+1)
plot(x$logFC,x$AveExpr,col=(x$adj.p.val < 0.05)+1)

```


### Functions
The above are all examples of built-in `functions` in R. A function in a programming language is like a function in math in that we pass variables into a function and it maps them to some output values. So here, when I use the function `max(x$logFC)`, I'm pulling the `logFC` column and getting its maximum value. 

It's useful to know that I can get help on any R function by using the question command in front of the function.  For example `?hist'. 

Later, you'll learn how to define your own functions. For operations that you might want to do over and over again, you can define your own function that takes as an input a value and returns some output. Here's a basic version...

```{r}

bring_umbrella <- function(weather){
  if(weather=="rainy"){
    return(TRUE)
  } else {
    return(FALSE)
  }
}

forecasts <- c("rainy","clear","rainy","snow","clear")

for(i in 1:length(forecasts)){
  if(bring_umbrella(forecasts[i])){
    print(paste("Day",i,"Bring umbrella!"))
  }
}

```

Or consider this more mathematical example... 

```{r}

quadratic <- function(x,a,b,c){
  return(c*(x-a)^2+b)
}


x<-seq(0,1,by=0.01)

y1<-quadratic(x,a=0,b=0,c=1) # standard classic quadratic with intercept and offset of zero
y2<-quadratic(x,a=0,b=1,c=-1) # shifted upside down up by 1

plot(x,y1)
plot(x,y2)

```

### Other versions of conditionals and loops

We can have more complex for versions of if/else

```{r}
z <- 5
if(z >= 5){
  print("z greater than or equal to 5")
}else if (z <= 2 ) {
  print("z less than or equal to 2")
} else {
  print("z between 2 and 5")
}
```

And `for` loops...
```{r}

# A "vector"/array of forecast values.  Each is a "character string"
forecasts <- c("rainy","clear","rainy","snow","clear")

for (i in forecasts)
  print(i)

i <-1
while(forecasts[i] != "snow"){
  print(paste("Day", i,"and still no snow."))
  i <- i + 1
}
print(paste("Day", i,"finally has snow."))
```


### Exercise

Write a program in R that uses a for loop to print the first 40 digits of a Fibonacci sequence.  Build it as a function `fibonacci(n)` that takes as it's input $n$, the lenght of the output sequence that you want to print. Share your program on slack when asked. 





### Why R's shortcuts work: vectorized operations

In the code above, `sum(x$Regulation == "Up")` replaced an entire for loop. This works because R is built around *vectorized operations*: when you apply a function or operator to a vector, R applies it to every element at once.

```{r}
x<-read.csv("Table_DEGs_GSE193677.csv")

logFC <- x$logFC

#logFC * 2                  # multiply every value by 2 — no loop needed
#logFC > 1                  # compare every value to 1 — returns a logical vector
sum(logFC > 1)             # count how many are TRUE (TRUE = 1, FALSE = 0)
mean(logFC > 1)            # proportion that are TRUE
plot(logFC, logFC*2)
```




Compare the loop version we wrote earlier with the vectorized version:

```{r}
# Loop version
ctr <- 0
for(i in 1:nrow(x)){
  if(x$logFC[i] > 1){ ctr <- ctr + 1 }
}
ctr

# Vectorized version
sum(x$logFC > 1)
```

Both give the same answer. The vectorized version is shorter, and on large datasets (millions of rows, common in genomics) it is also much faster.

The key mental shift is that instead of telling R what to do to each element one at a time, describe what to do to the whole vector at once. This is favored approach in R's design philosophy, and it's what makes R especially well-suited to data analysis.

## Simulation and randomness

Many biological processes are inherently *stochastic* — random: when does a molecule bind to a receptor? Which allele gets passed to an offspring? Does a cell divide or die? Programming lets us simulate these processes and study their behavior.

To use random numbers reproducibly, we first introduce `set.seed()`. Because each call to a random function gives different results, we fix the random number generator with a seed so the same code always produces the same output — an essential habit for reproducible science.

```{r}
set.seed(42)    # any integer; your collaborator running this code will get identical results
```

Now let's simulate a random walk: a sequence of steps where each step is randomly +1 or -1, chosen with equal probability. Random walks are a model for Brownian motion (diffusion of a molecule in a cell), genetic drift (random fluctuation of allele frequencies), and many other biological phenomena.

```{r}
# Initialize random number seed, number of steps, and position
set.seed(42)
n_steps <- 200
position <- rep(0,n_steps) 

# Loop through steps
for( i in 2:n_steps){
  # Sample a step up or down with equal probability
  one_step <- sample(c(-1, 1), size = 1, replace = TRUE)
  # Increment position
  position[i] <- position[i-1] + one_step
}

# Plot
plot(1:n_steps, position, type = "l",
     xlab = "Step", ylab = "Position",
     main = "A random walk")
abline(h = 0, col = "gray", lty = 2)
```
Even though each step is 50/50, the walk drifts away from zero. Change the seed and re-run to see how variable the outcome is. This variability is exactly what makes stochastic models interesting — and challenging to reason about without simulation.

Exercise: Use a for loop to simulate 5 independent random walks and plot them all on the same axes using plot() for the first and lines() for the rest. What do you observe about how far the walks spread as the number of steps grows?

In the Defensive Programming tutorial you'll write stochastic simulations of population dynamics. The habits to build now are: always set.seed() before a simulation, and always plot the output to sanity-check it before drawing conclusions.


### Data frames: R's central data structure

The object `x` that `read.csv` created is called a **data frame** - this is R's data strucutre that maps closely to the familiar structure of data table that you've seen in a spreadsheet. Its rules are simple:

- each **row** is one observation (here, one gene)
- each **column** is one variable (gene symbol, logFC, p-value, …)
- every column is a typed vector — all entries in a column share the same type

```{r}
x <- read.csv("Table_DEGs_GSE193677.csv")

nrow(x)          # how many genes?
ncol(x)          # how many columns?
colnames(x)      # what are they called?
str(x)           # compact summary: column name, type, and first few values
```
str() is one of the most useful functions in R for understanding a new dataset — run it on every table you encounter.

We can pull out pieces of the data frame in several ways:
```{r}
x$logFC          # an entire column, by name — returns a vector
x[1, ]           # the first row, all columns
print(x[1, "GeneSymbol"])  # one cell: row 1, GeneSymbol column
```
And we can filter rows using a logical condition — this pattern is extremely common:
```{r}
up_genes <- x[x$Regulation == "Up", ]   # keep only rows where Regulation is "Up"
nrow(up_genes)
head(up_genes)
```
This gives the same result as the for-loop counter we wrote before — but it's one line. This is an example of how a specialized language like R simplifies our work.  The Data Visualization will show you even more powerful ways to filter and reshape data frames using the dplyr package.

(Also note that the `tibble` is closely related to a `data.frame` and is an update version of the `data.frame` data struture)

### Exercise

Use real forecast data to make a predictor of whether to bring an umbrella tomorrow.

```{r}
# Install from CRAN or GitHub
install.packages("weathR")
library(weathR)
```

```{r}
# Fetch the forecast using latitude and longitude coordinates
us_forecast <- weathR::point_tomorrow(lat = 44.5994, lon = -72.6352)

# View data (returns a tidy simple features object where possible)
head(us_forecast)
```




